---
title: "Cholera Trigger Analysis"
date: "`r Sys.Date()`"
output: html_notebook
---

This document contains analysis used to create an alert system using the WHO bulletins for cholera. It is done by country and checks for common indicators of an uptick in cases.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999)
min_cfr <- 0
min_cases <- 500
```


```{r load-libraries, echo=FALSE}
box::use(
  readr,
  dplyr,
  purrr,
  lubridate,
  janitor,
  stringr,
  zoo,
  ggplot2,
  gghdx,
  gt
)
box::use(src/location_codes)
gghdx$gghdx()
iso3_to_plot <- c("CMR", "COD", "ETH", "MOZ", "NGA", "SSD") #iso3_levels

```

### Load Data

There are 3 columns with start of reporting period. It seems these columns are reporting on the same value and if one has a value, the rest are NA.

```{r load-data, echo=FALSE}
df_raw_link <- "https://raw.githubusercontent.com/CBPFGMS/pfbi-data/refs/heads/main/final_data_for_powerbi_with_kpi.csv"
df_raw <- readr$read_csv(df_raw_link)

df_cerf <- readr$read_csv(file.path(Sys.getenv("AA_DATA_DIR"), "public", "raw", "glb", "cerf_allocations", "250509_CERF cholera allocations(Data).csv"))
cerf_cholera <- df_cerf[!(is.na(df_cerf$`Application Code`) | df_cerf$`Application Code` == ""), ] |>
  dplyr$mutate(countryCode = stringr$str_extract(`Application Code`, "(?<=-)[A-Z]{3}(?=-)"))
df_raw |> 
  head(5) |>
  gt$gt()

```

### Clean and Prepare Data

```{r clean-data, echo=FALSE}
df_clean <- df_raw |>
  # Clean column names to snake_case using janitor
  janitor$clean_names() |>

  # Filter for rows where the event contains 'cholera' but exclude other similar diseases
  dplyr$filter(
    stringr$str_detect(tolower(event), "cholera") &
      !stringr$str_detect(tolower(event), "intestinal|bacterial|shigellosis|salmonellosis")
  ) |>

  # Select and rename key columns
  dplyr$transmute(
    iso3 = location_codes$names_to_iso3(country),  # Convert country names to ISO3 codes
    event,

    # Consolidate multiple possible start date fields
    start_date_raw = dplyr$case_when(
      !is.na(start_of_reporting_period) ~ start_of_reporting_period,
      !is.na(start_of_reporting_period_2) ~ start_of_reporting_period_2,
      !is.na(start_of_reporting_period_3) ~ start_of_reporting_period_3
    ),

    date = as.Date(week_date),  # Parse the week_date column to Date format
    #week_date_start = date - lubridate$wday(date, week_start = 1) + 1,  # Calculate the start of the week

    # Parse start_date using appropriate format based on date patterns or time
    start_date = dplyr$case_when(
      stringr$str_detect(start_date_raw, "[0-9]{1,2}[-|//][A-Za-z]{3}") ~ lubridate$dmy(start_date_raw, quiet = TRUE),
      stringr$str_detect(start_date_raw, "^[A-Za-z]{3}") ~ lubridate$mdy(start_date_raw, quiet = TRUE),
      date >= "2023-09-25" ~ lubridate$mdy(start_date_raw, quiet = TRUE),
      date < "2023-09-25" ~ lubridate$dmy(start_date_raw, quiet = TRUE)
    ),

    cholera_cases = readr$parse_number(as.character(total_cases)),  # Extract numeric case count
    cholera_cfr = readr$parse_number(as.character(cfr)),
    deaths = readr$parse_number(as.character(deaths))

  ) |>

  # Drop the raw start_date column
  dplyr$select(-start_date_raw) |>

  # Group by ISO3 and date for aggregation
  dplyr$group_by(iso3, date) |>

  # Summarize by collapsing multiple event descriptions and summing cases
  dplyr$summarize(
    event = paste(unique(event), collapse = "; "),
    start_date = min(start_date),
    cholera_cases = sum(cholera_cases),
    cholera_cfr = sum(cholera_cfr),
    deaths = sum(deaths),
    .groups = "drop"
  ) |>

  # Re-group by country
  dplyr$group_by(iso3) |>

  # Sort by date within each group
  dplyr$arrange(date, .by_group = TRUE) |>

  # Interpolate missing case values using linear approximation
  dplyr$mutate(
    cholera_cases = zoo$na.approx(cholera_cases, na.rm = FALSE)
  ) |>
  # Fill in missing dates with NA values for cholera cases
  # Cleaning up values shown as decimals

  # Remove grouping
  dplyr$ungroup()
df_clean |>
  head(5) |>
  gt$gt()
```

First, checking if values tend to be increasing (cumulative) for each country
Most of the values are greater than or equal to the one prior to it. (lowest being close to 90%), giving a strong indication of cumulative number of cases.

```{r increase-check, echo=FALSE}
df_rising_check <- df_clean |>
  dplyr$group_by(iso3) |>
  dplyr$arrange(iso3, date) |>
  dplyr$mutate(
    rising_cases = cholera_cases >= dplyr$lag(cholera_cases, 1, default = NA),
  ) |>
  dplyr$summarise(mean(rising_cases, na.rm = TRUE))
df_rising_check |>
  head(5) |>
  gt$gt()
```


### Additional Data Cleaning

 - These weekdays for the date change
 - these dates change depending on the year and weeks can change even within the year
 - they tend to be between Mon-Wed
 - tried to fill in values and seems there is an issue with the definition of weeks
 - Also to replace values of 1 and correcting those with decimals especially those around other larger values
 - Fill in the missing weeks

```{r additional-clean, echo=FALSE}
decimal_values <- df_clean |>
  dplyr$filter(cholera_cases %% 1 != 0)
small_values <- df_clean |>
  dplyr$filter(cholera_cases < 200)
df_prepped <- df_clean |>
  dplyr$group_by(iso3) |>
  dplyr$mutate(
    original_value = cholera_cases,
    # if previous value is more than 3 weeks ago, do not change small values
    prev_date_diff = as.numeric(date - dplyr$lag(date, 1, default = NA)),
    # Create the previous and next value columns
    # Get the previous values
    prev_val = pmax(
      dplyr$lag(cholera_cases, 1, default = NA),
      dplyr$lag(cholera_cases, 2, default = NA),
      dplyr$lag(cholera_cases, 3, default = NA),
      dplyr$lag(cholera_cases, 4, default = NA),
      dplyr$lag(cholera_cases, 5, default = NA),
      na.rm = TRUE
    ),
    # Get the next values
    next_val = pmax(
      dplyr$lead(cholera_cases, 1, default = NA),
      dplyr$lead(cholera_cases, 2, default = NA),
      dplyr$lead(cholera_cases, 3, default = NA),
      dplyr$lead(cholera_cases, 4, default = NA),
      dplyr$lead(cholera_cases, 5, default = NA),
      na.rm = TRUE
    ),
    # Calculate the reference value as the larger of the previous and next values
    # Calculate the scaled cholera_cases (multiplying by 10, 100, and 1000)
    scaled_1 = cholera_cases * 1,
    scaled_10 = cholera_cases * 10,
    scaled_100 = cholera_cases * 100,
    scaled_1000 = cholera_cases * 1000,
    scaled_10000 = cholera_cases * 10000,
    # Calculate the absolute differences between scaled values and the reference
    diffs_prev_1 = abs(scaled_1 - prev_val),
    diffs_prev_10 = abs(scaled_10 - prev_val),
    diffs_prev_100 = abs(scaled_100 - prev_val),
    diffs_prev_1000 = abs(scaled_1000 - prev_val),
    diffs_prev_10000 = abs(scaled_10000 - prev_val),
    diffs_next_1 = abs(scaled_1 - next_val),
    diffs_next_10 = abs(scaled_10 - next_val),
    diffs_next_100 = abs(scaled_100 - next_val),
    diffs_next_1000 = abs(scaled_1000 - next_val),
    diffs_next_10000 = abs(scaled_10000 - next_val),
    # Find the minimum difference for each row
    min_diff = pmin(diffs_prev_1, diffs_prev_10, diffs_prev_100, diffs_prev_1000,
                    diffs_prev_10000, diffs_next_1, diffs_next_10, diffs_next_100,
                    diffs_next_1000, diffs_next_10000, na.rm = TRUE),
    # Find the best match by selecting the scaled value with the smallest difference
    best_match = floor(dplyr$case_when(
      min_diff == diffs_prev_1 ~ scaled_1,
      min_diff == diffs_prev_10 ~ scaled_10,
      min_diff == diffs_prev_100 ~ scaled_100,
      min_diff == diffs_prev_1000 ~ scaled_1000,
      min_diff == diffs_prev_10000 ~ scaled_10000,
      min_diff == diffs_next_1 ~ scaled_1,
      min_diff == diffs_next_10 ~ scaled_10,
      min_diff == diffs_next_100 ~ scaled_100,
      min_diff == diffs_next_1000 ~ scaled_1000,
      min_diff == diffs_next_10000 ~ scaled_10000,
      TRUE ~ cholera_cases
    )),
    best_match_class = ceiling(best_match / cholera_cases),
    ref = dplyr$case_when(
      is.na(prev_val) & is.na(next_val) ~ NA_real_,
      is.na(prev_val) ~ next_val,
      is.na(next_val) ~ prev_val,
      abs(prev_val - best_match) <= abs(next_val - best_match) ~ prev_val,
      TRUE ~ next_val
    ),
    # Apply the scaling logic: update cholera_cases based on the best match
    cholera_cases = dplyr$case_when(
      # prev_value == next value == current value, then keep current value
      (dplyr$lag(cholera_cases, 1) == dplyr$lead(cholera_cases, 1) &
         dplyr$lag(cholera_cases, 1) == cholera_cases) ~ cholera_cases,
      # If the value is non-integer and ref exists, check min_diff against best_match_class
      ((cholera_cases %% 1 != 0 & !is.na(ref))) ~ {
        dplyr$if_else((min_diff < 1 | min_diff < best_match_class |
                         best_match_class >= 100), best_match, floor(cholera_cases))
      },
      # If cholera_cases is small (< 200), and strong match to prev/next
      cholera_cases < 200 & !is.na(ref) ~ {
        dplyr$if_else((prev_val == next_val | min_diff < best_match_class) &
                        (!is.na(prev_val) & !is.na(next_val)) &
                        (prev_date_diff <= 28), ref, cholera_cases)
      },
      # If min_diff is smaller than best_match_class, update
      min_diff < best_match_class & !is.na(ref) ~ {
        best_match
      },
      TRUE ~ floor(cholera_cases)
    )
  ) |>
  dplyr$select(iso3, date, event, start_date, cholera_cases, cholera_cfr, deaths, prev_date_diff)
df_prepped |>
  head(5) |>
  gt$gt()
```

### Weekly Check
Quick check on how many weeks per country and year in dataset
```{r weeks-per-country, echo=FALSE}
df_weeks_per_country <- df_prepped |>
  dplyr$mutate(year = lubridate$year(date)) |>
  dplyr$group_by(iso3, year) |>
  dplyr$summarize(weeks = dplyr$n(), .groups = "drop") |>
  dplyr$ungroup() |>
  dplyr$group_by(iso3) |>
  dplyr$summarize(min_weeks = min(weeks), max_weeks = max(weeks), mean_weeks = mean(weeks), .groups = "drop")
df_weeks_per_country |>
  head(5) |>
  gt$gt()
```

Check for countries where dates are less than 7 days apart

```{r check-date-diffs, echo=FALSE}
df_prepped |>
  dplyr$filter(prev_date_diff < 7) |>
  dplyr$group_by(iso3) |>
  dplyr$summarise(count = dplyr$n()) |>
  head(5) |>
  gt$gt()
```

### Global Cholera Trigger Options:

1. Use the DRC trigger:
   - Level 3 alert sustained for 3 weeks
   - Level 3 defined as either:
       a) 4x increase compared to baseline, or
       b) Value above the 99th percentile

2. Use the old cumulative methodology (as included by Seth in the Signals system)

3. Use the Signals methodology:
   - Alert when current value is higher than any point in the past 12 months
4. Use a 3-year rolling average of weekly increases

Exploration of using the CFR or the Number of Deaths as well.

CERF Allocation Years

```{r find-nearest-country, echo=FALSE}
find_nearest_country <- function(event_df, allocation_df) {
  event_df |>
    dplyr$group_by(iso3) |>  # Group by country
    dplyr$rowwise() |>
    dplyr$mutate(
      # pull only the allocation dates for this same iso3
      country_alloc_dates = list(allocation_df$dateUSGSignature[allocation_df$countryCode == iso3]),
      # pick the one closest to this event’s date
      # if no dates, give NA; otherwise pick the one with the smallest abs diff
      nearest_allocation_date = if (length(country_alloc_dates) == 0) {
        NA
      } else {
        country_alloc_dates[which.min(abs(difftime(date, country_alloc_dates, units = "days")))]
      },
      # compute signed or absolute difference as you wish
      time_diff_days = as.numeric(difftime(nearest_allocation_date, date, units = "days"))
    ) |>
    dplyr$ungroup()
}
allocation_years <- cerf_cholera |>
  #dplyr$filter(!is.na(dateUSGSignature)) |>
  #dplyr$mutate(year = lubridate$year(dateUSGSignature)) |>
  dplyr$distinct(countryCode, Year) |>
  dplyr$filter(Year >= lubridate$year(min(df_prepped$date, na.rm = TRUE))) |>
  dplyr$rename(iso3 = countryCode)
allocation_years |>
  head(5) |>
  gt$gt()
```

## Methodology 1
Assumption: Values stay the same across weeks without data until reported otherwise and also making the data noncumulative.
```{r plot-alerts-by-year-and-level, echo=FALSE, fig.width=10, fig.height=6}
df_1_all <- df_prepped |>
  dplyr$group_by(iso3) |>
  dplyr$mutate(
    weekly_increase = dplyr$if_else(prev_date_diff <= 28,
                                    cholera_cases - dplyr$lag(cholera_cases, 1),
                                    NA_real_),
    p95 = quantile(weekly_increase[weekly_increase > 0], probs = 0.95, na.rm = TRUE),
    p975 = quantile(weekly_increase[weekly_increase > 0], probs = 0.975, na.rm = TRUE),
    p99 = quantile(weekly_increase[weekly_increase > 0], probs = 0.99, na.rm = TRUE),
    year = lubridate$year(date),
    alert_level = dplyr$case_when(
      weekly_increase >= min_cases & weekly_increase >= p99 & cholera_cfr > min_cfr ~ "p99",
      weekly_increase >= min_cases & weekly_increase >= p975 ~ "p975",
      weekly_increase >= min_cases & weekly_increase >= p95 ~ "p95",
      TRUE ~ "none"
    )
  )
df_1 <- df_1_all |>
  dplyr$ungroup() |>
  dplyr$mutate(fold_increase = weekly_increase / dplyr$lag(weekly_increase, 1, default = NA))

# defining levels, 95 as baseline, 97.5 as watch and 99 as alert
# testing how often each is reached
# Summarize the number of alerts by country, year, and alert level
alerts_by_year_and_level <- df_1 |>
  dplyr$count(iso3, year, alert_level, name = "num_alerts") |>
  dplyr$filter(alert_level != "none")

# Plot the heatmap of the number of alerts by country, year, and alert level
ggplot2$ggplot(alerts_by_year_and_level, ggplot2$aes(x = year, y = iso3, fill = num_alerts)) +
  ggplot2$geom_tile() +  # Use tiles to create a heatmap
  ggplot2$geom_tile(
    data = allocation_years,
    mapping = ggplot2$aes(x = Year, y = iso3),
    fill = NA,
    color = "darkgreen",
    linewidth = 0.55
  ) +
  ggplot2$geom_text(ggplot2$aes(label = num_alerts), color = "black", size = 3) +
  ggplot2$labs(
    title = "Number of Alerts by Country, Year, and Alert Level",
    x = "Year",
    y = "Country",
    fill = "Number of Alerts",
    caption = "Green outlines indicate years with a CERF allocation"
  ) +
  ggplot2$facet_wrap(~alert_level) +  # Facet by alert level to create separate tiles for each level
  ggplot2$scale_fill_gradient2(low = "white", high = "red", limits = c(0, 10),
                               breaks = function(x) x[x %% 1 == 0],
                               labels = function(x) x[x %% 1 == 0]) +  # Adjust labels
  ggplot2$scale_y_discrete(limits = rev(levels(factor(alerts_by_year_and_level$iso3)))) +
  ggplot2$scale_x_continuous(breaks = unique(alerts_by_year_and_level$year))

```

#### Comparison with CERF allocations
which alert level correlates with a CERF allocation (within 30 days before an alert or 90 days after)
```{r compare-with-cerf, echo=FALSE}
df_1_cerf <- find_nearest_country(event_df = df_1, allocation_df = cerf_cholera)
df_summary <- df_1_cerf |>
  # baseline of 1000
  dplyr$mutate(
    n_window = time_diff_days >= -30 & time_diff_days <= 90
  ) |>
  dplyr$filter(weekly_increase >= 1000 & alert_level != "none" & n_window)
df_summary |>
  dplyr$group_by(alert_level) |>
  dplyr$summarise(
    n = dplyr$n(),
    n_window = sum(n_window, na.rm = TRUE),
    prop_window = n_window / n,
    .groups = "drop"
  )
df_summary |>
  head(5) |>
  gt$gt()
```

## Methodology 2
This method should use cumulative values.
```{r methodology-2, echo=FALSE}
df_2_all <- df_prepped |>
  dplyr$arrange(iso3, date) |>
  dplyr$mutate(
    weekly_increase = dplyr$if_else(prev_date_diff <= 28, cholera_cases - dplyr$lag(cholera_cases, 1), NA_real_),
    level = dplyr$case_when(
      weekly_increase >= 5000 & dplyr$lag(weekly_increase, default = 0) < 5000 ~ "Level 2",
      weekly_increase >= 1000 & dplyr$lag(weekly_increase, default = 0) < 1000 ~ "Level 1",
      TRUE ~ NA_character_
    )
  )
df_2_a <- df_2_all |>
  dplyr$filter(!is.na(level))

# how many alerts are generated by country
# Calculate the number of alerts per year for each country
alerts_by_year <- df_2_a |>
  dplyr$mutate(year = lubridate$year(date)) |>
  dplyr$count(iso3, year, level, name = "num_alerts") |>
  dplyr$arrange(iso3)

# Plot a box plot of the number of alerts per year by country
# Create a heatmap to visualize the number of alerts by country and year
ggplot2$ggplot(alerts_by_year, ggplot2$aes(x = year, y = iso3, fill = num_alerts)) +
  ggplot2$geom_tile() +
  ggplot2$geom_tile(
    data = allocation_years,
    mapping = ggplot2$aes(x = Year, y = iso3),
    fill = NA,
    color = "darkgreen",
    linewidth = 0.55
  ) +
  ggplot2$geom_text(ggplot2$aes(label = num_alerts), color = "black", size = 3) +
  ggplot2$labs(
    title = "Number of Alerts by Country, Year, and Alert Level",
    x = "Year",
    y = "Country",
    fill = "Number of Alerts",
    caption = "Green outlines indicate years with a CERF allocation"
  ) +
  ggplot2$facet_wrap(~level) +
  ggplot2$scale_fill_gradient(low = "white", high = "red", limits = c(0, 10),
                              breaks = function(x) x[x %% 1 == 0],
                              labels = function(x) x[x %% 1 == 0]) +
  ggplot2$scale_y_discrete(limits = rev(levels(factor(alerts_by_year$iso3)))) +
  ggplot2$scale_x_continuous(breaks = unique(alerts_by_year$year))
```

Comparison with CERF allocations
```{r compare-with-cerf-methodology-2, echo=FALSE}
df_2a_cerf <- find_nearest_country(event_df = df_2_a, allocation_df = cerf_cholera)
df_2a_cerf |>
  dplyr$summarise(
    n = dplyr$n(),
    n_window = sum(time_diff_days >= -60 & time_diff_days <= 90, na.rm = TRUE),
    prop_window = n_window / n,
  ) |>
  head(5) |>
  gt$gt()
```

## Methodology 3
Remove zeros from weekly increases and adding check for no signals for the next 3 months
```{r methodology-3, echo=FALSE}
df_3_all <- df_prepped |>
  dplyr$group_by(iso3) |>
  dplyr$mutate(
    weekly_increase = dplyr$if_else(prev_date_diff <= 28, cholera_cases - dplyr$lag(cholera_cases, 1), 0),
    alert_initial = purrr$map_lgl(date, ~{
      window_start <- .x - months(12)
      window_vals <- weekly_increase[date > window_start & date <= .x]

      if (all(is.na(window_vals))) {
        FALSE
      } else {
        current_val <- weekly_increase[date == .x]
        cholera_cfr <- cholera_cfr[date == .x]
        max_val <- max(window_vals, na.rm = TRUE)
        (current_val == max_val) && (current_val >= min_cases) && (cholera_cfr > min_cfr)
      }
    }),
    alert = purrr$map2_lgl(date, seq_along(date), ~{
      if (!alert_initial[.y] || weekly_increase[.y] < min_cases) return(FALSE)

      last_alert_date <- max(date[alert_initial & date < .x], na.rm = TRUE)

      if (!is.finite(last_alert_date)) {
        TRUE
      } else {
        difftime(.x, last_alert_date, units = "days") > 90
      }
    })
  )
df_3 <- df_3_all |>
  dplyr$filter(alert)


# comparing with CERF allocations
df_3_cerf <- find_nearest_country(event_df = df_3, allocation_df = cerf_cholera)
# Aggregating the data for CERF allocation windows
df_3_aggregated <- df_3_cerf |>
  dplyr$group_by(iso3, year = lubridate$year(date)) |>
  dplyr$count(alert, name = "num_alerts") |>
  dplyr$ungroup()


# plotting the heat map
ggplot2$ggplot(df_3_aggregated, ggplot2$aes(x = year, y = iso3, fill = num_alerts)) +
  ggplot2$geom_tile() +
  ggplot2$geom_tile(
    data = allocation_years,
    mapping = ggplot2$aes(x = Year, y = iso3),
    fill = NA,
    color = "darkgreen",
    linewidth = 0.55
  ) +
  ggplot2$geom_text(ggplot2$aes(label = num_alerts), color = "black", size = 3) +
  ggplot2$labs(
    title = "Number of Alerts by Country and Year",
    x = "Year",
    y = "Country",
    fill = "Number of Alerts",
    caption = "Green outlines indicate years with a CERF allocation"
  ) +
  ggplot2$scale_fill_gradient(low = "white", high = "red", limits = c(0, 10),
                              breaks = function(x) x[x %% 1 == 0],
                              labels = function(x) x[x %% 1 == 0]) +
  ggplot2$scale_y_discrete(limits = rev(levels(factor(df_3_aggregated$iso3)))) +
  ggplot2$scale_x_continuous(breaks = unique(df_3_aggregated$year))
```

## Methodology 4
Create a 3-year rolling average of weekly increases with a minimum of 500 cases. 

```{r rolling-average, echo=FALSE, warning=FALSE}
df_4 <- df_prepped |>
  dplyr$group_by(iso3) |>
  dplyr$group_modify(~ {
    df <- .x
    df$weekly_increase <- dplyr$if_else(
      df$prev_date_diff <= 28,
      df$cholera_cases - dplyr$lag(df$cholera_cases, 1),
      0
    )
    
    valid_cases <- dplyr$if_else(df$weekly_increase > 0, df$weekly_increase, NA_real_)
    n_valid <- sum(!is.na(valid_cases))
    
    if (n_valid >= 156) {
      roll <- zoo$rollmean(valid_cases, k = 156, fill = NA, align = "right")
      first_val <- roll[which(!is.na(roll))[1]]
      roll[is.na(roll)] <- first_val
    } else {
      roll <- rep(mean(valid_cases, na.rm = TRUE), length(valid_cases))
    }
    
    df$roll_3yr <- roll
    df$alert <- df$weekly_increase >= df$roll_3yr & df$weekly_increase >= min_cases
    df$year <- lubridate$year(df$date)
    df
  }) |>
  dplyr$ungroup()

# Aggregating the data for CERF allocation windows
df_4_aggregated <- df_4 |>
  dplyr$count(iso3, year, alert, name = "num_alerts") |>
  dplyr$filter(alert)

ggplot2$ggplot(df_4_aggregated, ggplot2$aes(x = year, y = iso3, fill = num_alerts)) +
  ggplot2$geom_tile() +
  ggplot2$geom_tile(
    data = allocation_years,
    mapping = ggplot2$aes(x = Year, y = iso3),
    fill = NA,
    color = "darkgreen",
    linewidth = 0.55
  ) +
  ggplot2$geom_text(ggplot2$aes(label = num_alerts), color = "black", size = 3) +
  ggplot2$labs(
    title = "Number of Alerts by Country and Year",
    x = "Year",
    y = "Country",
    fill = "Number of Alerts",
    caption = "Green outlines indicate years with a CERF allocation"
  ) +
  ggplot2$scale_fill_gradient(low = "white", high = "red", limits = c(0, 12),
                              breaks = function(x) x[x %% 1 == 0],
                              labels = function(x) x[x %% 1 == 0]) +
  ggplot2$scale_y_discrete(limits = rev(levels(factor(df_4_aggregated$iso3)))) +
  ggplot2$scale_x_continuous(breaks = unique(df_4_aggregated$year))
```
Timelines for Method 4.
Challenges with method 4:

 - We do not always have adequate data over 3 years for most countries and it is not always consecutive weeks. There are gaps in the weeks.
 - The rolling average results in multiple alerts even when not needed.

```{r rolling-average-tl, echo=FALSE, warning=FALSE}
purrr$walk(iso3_to_plot, ~ {
  cerf_dates <- cerf_cholera$final_date[cerf_cholera$countryCode == .x]
  df_country <- df_4[df_4$iso3 == .x, ]
  print({
    ggplot2$ggplot(df_country, ggplot2$aes(x = date, y = weekly_increase)) +
      ggplot2$geom_line(size = 0.6, color = "grey40") +
      ggplot2$geom_line(ggplot2$aes(y = roll_3yr), color = "green4", size = 0.5, na.rm = TRUE) +
      ggplot2$geom_point(data = subset(df_country, alert == TRUE),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "orange", size = 1.8) +
      ggplot2$geom_vline(xintercept = as.numeric(cerf_dates), color = "green", size = 0.7) +
      ggplot2$geom_hline(yintercept = 500, color = "darkblue", linetype = "dashed", show.legend = FALSE) +
      ggplot2$annotate("text", x = as.Date("2025-08-01"), y = 500, label = "500",
                       color = "darkblue", hjust = 0, vjust = -0.5) +
      ggplot2$scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
      ggplot2$labs(
        title = paste("Weekly Cholera Case Increase –", .x),
        x = "Date",
        y = "Weekly Increase in Cases",
        caption ="
        ● Orange Dot: Rolling average alert
        ● Green line: CERF allocation date
        ● Dashed line: 500 cases threshold
        ● Green curve: 3-year rolling average"
      ) +
      gghdx$gghdx() +
      ggplot2$theme(
        plot.title = ggplot2$element_text(face = "bold", size = 14),
        axis.title = ggplot2$element_text(size = 11),
        axis.text = ggplot2$element_text(size = 10),
        plot.caption = ggplot2$element_text(size = 9, hjust = 0),
        panel.grid.minor = ggplot2$element_blank()
      )
  })
})

```


### Time Series of Cases by Country
A time series of cases by country showing the alert levels

```{r plot-setup, echo=FALSE}

df_joined <- dplyr$full_join(df_1, df_2_a,
                             by = c("iso3", "date"),
                             suffix = c("_df1", "_df2a")) |>
  dplyr$full_join(df_3, by = c("iso3", "date"))
head(df_joined)
combined_df <- df_1_all |>
  dplyr$left_join(df_2_all, by = c("iso3", "date")) |>
  dplyr$left_join(df_3_all, by = c("iso3", "date"))
cerf_cholera <- cerf_cholera |>
  dplyr$mutate(final_date = as.Date(dplyr$coalesce(`Date of Initial Contact`,
                                                   `Date of ERC Endorsement`,
                                                   `First Project Approved`,
                                                   `Date of Earliest Project Start`),
                                    format = "%d/%m/%Y"))

```

```{r time-series-alerts, echo=FALSE, eval=FALSE}
purrr$walk(iso3_levels, ~ {
  print({
    ggplot2$ggplot(combined_df[combined_df$iso3 == .x, ], ggplot2$aes(x = date, y = weekly_increase)) +
      ggplot2$geom_line() +
      ggplot2$geom_point(data = subset(combined_df, iso3 == .x & alert_level == "p99"),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "red", size = 2) +
      ggplot2$geom_point(data = subset(combined_df, iso3 == .x & level == "Level 2"),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "orange", size = 2) +
      ggplot2$geom_point(data = subset(combined_df, iso3 == .x & alert == TRUE),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "#0F52BA", size = 2) +
      ggplot2$geom_hline(ggplot2$aes(yintercept = p99), color = "red", linetype = "dashed", show.legend = FALSE) +
      ggplot2$geom_hline(yintercept = 1000, color = "darkgreen", linetype = "dashed", show.legend = FALSE) +
      ggplot2$geom_hline(yintercept = 5000, color = "orange", linetype = "dashed", show.legend = FALSE) +
      ggplot2$annotate("text", x = as.Date("2021-01-01"), y = 1000, label = "1000",
                       color = "darkgreen", hjust = 0, vjust = -0.5) +
      ggplot2$annotate("text", x = as.Date("2022-01-01"), y = 5000, label = "5000",
                       color = "orange", hjust = 0, vjust = -0.5) +
      ggplot2$geom_text(ggplot2$aes(x = as.Date("2020-01-01"), y = p99, label = "p99"),
                        color = "red", hjust = 0, vjust = -0.5) +
      ggplot2$labs(
        title = paste("Weekly Increase in Cholera Cases:", .x),
        x = "Date",
        y = "Weekly Increase",
        caption = "red dots show alerts for method 1
        orange dots show alerts for method 2 \nblue dots show an alert for method 3"
      ) +
      gghdx$gghdx()
  })
})
```

## Final proposal
    - Watch alert: highest value in the last 12 months
    - Warning alert: 99th percentile
```{r final-proposal, echo=FALSE}
p99_heat <- df_1 |>
  dplyr$filter(alert_level == "p99") |>
  dplyr$mutate(year = lubridate$year(date), method = "Warning Alert: 99th percentile") |>
  dplyr$count(iso3, year, method, name = "num_alerts")

df_3_heat <- df_3_aggregated |>
  dplyr$mutate(method = "Watch Alert: Highest value") |>
  dplyr$rename(num_alerts = num_alerts)

# Combine
heat_df <- dplyr$bind_rows(df_3_heat, p99_heat) |>
  dplyr$mutate(method = factor(method,
                               levels = c("Watch Alert: Highest value",
                                          "Warning Alert: 99th percentile")))

ggplot2$ggplot(heat_df, ggplot2$aes(x = year, y = iso3, fill = num_alerts)) +
  ggplot2$geom_tile() +
  ggplot2$geom_tile(
    data = allocation_years,
    mapping = ggplot2$aes(x = Year, y = iso3),
    fill = NA,
    color = "darkgreen",
    linewidth = 0.55
  ) +
  ggplot2$geom_text(ggplot2$aes(label = num_alerts), color = "black", size = 3) +
  ggplot2$facet_wrap(~ method) +
  ggplot2$scale_fill_gradient(low = "white", high = "red", limits = c(0, 10),
                              breaks = function(x) x[x %% 1 == 0],
                              labels = function(x) x[x %% 1 == 0]) +
  ggplot2$scale_y_discrete(limits = rev(levels(factor(heat_df$iso3)))) +
  ggplot2$scale_x_continuous(breaks = unique(heat_df$year)) +
  ggplot2$labs(
    title = paste0("Number of Alerts per Country and Year (min CFR ", min_cfr,"%)"),
    x = "Year",
    y = "Country",
    fill = "Number of Alerts",
    caption = "Green borders = CERF allocation years"
  )
```

#### Plotting for a specific country

```{r specific-country-plots, echo=FALSE}

purrr$walk(iso3_to_plot, ~ {
  cerf_dates <- cerf_cholera$final_date[cerf_cholera$countryCode == .x]
  df_country <- combined_df[combined_df$iso3 == .x, ]
  print({
    ggplot2$ggplot(df_country, ggplot2$aes(x = date, y = weekly_increase)) +
      ggplot2$geom_line(size=0.6, color = "grey40") +
      ggplot2$geom_point(data = subset(df_country, alert == TRUE),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "orange", size = 1.8) +
      ggplot2$geom_point(data = subset(df_country, alert_level == "p99"),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "red", size = 1.8) +
      ggplot2$geom_vline(xintercept = as.numeric(cerf_dates), color = "green", size=0.7) +
      ggplot2$geom_hline(ggplot2$aes(yintercept = p99), color = "red", linetype = "dashed", show.legend = FALSE) +
      ggplot2$geom_hline(yintercept = 500, color = "darkblue", linetype = "dashed", show.legend = FALSE) +
      ggplot2$annotate("text", x = as.Date("2025-08-01"), y = 500, label = "500",
                       color = "darkblue", hjust = 0, vjust = -0.5) +
      ggplot2$geom_text(ggplot2$aes(x = as.Date("2020-01-01"), y = p99, label = "99th percentile"),
                        color = "red", hjust = 0, vjust = -0.5) +
      ggplot2$scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
      ggplot2$labs(
        title = paste("Weekly Cholera Case Increase –", .x),
        x = "Date",
        y = "Weekly Increase in Cases",
        caption ="
        ● Orange Dot: Watch alert
        ● Red Dot: Warning alert
        ● Green line: CERF allocation date
        ● Dashed lines: 500 cases & 99th percentile thresholds"
      ) +
      gghdx$gghdx() +
      ggplot2$theme(
        plot.title = ggplot2$element_text(face = "bold", size = 14),
        axis.title = ggplot2$element_text(size = 11),
        axis.text = ggplot2$element_text(size = 10),
        plot.caption = ggplot2$element_text(size = 9, hjust = 0),
        panel.grid.minor = ggplot2$element_blank()
      )
  })
})
```


#### Plots with CFR

```{r specific-country-plots-cfr, echo=FALSE}
purrr$walk(iso3_to_plot, ~ {
  cerf_dates <- cerf_cholera$final_date[cerf_cholera$countryCode == .x]
  df_country <- combined_df[combined_df$iso3 == .x, ]
  max_y <- max(df_country$weekly_increase, na.rm = TRUE)
  scale_factor <- max_y / 8
  # print(paste("Scale Factor: ", ceiling(scale_factor)))
  print({
    ggplot2$ggplot(df_country, ggplot2$aes(x = date, y = weekly_increase)) +
      ggplot2$geom_line(size=0.6, color = "grey40") +
      ggplot2$geom_point(data = subset(df_country, alert == TRUE),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "orange", size = 1.8) +
      ggplot2$geom_point(data = subset(df_country, alert_level == "p99"),
                         ggplot2$aes(x = date, y = weekly_increase),
                         color = "red", size = 1.8) +
      ggplot2$geom_vline(xintercept = as.numeric(cerf_dates), color = "green", size=0.7) +
      ggplot2$geom_line(ggplot2$aes(y = cholera_cfr * scale_factor), color = "maroon", linetype = "dashed") +
      ggplot2$scale_y_continuous(
        name = "Weekly Increase in Cases",
        sec.axis = ggplot2$sec_axis(~ . / scale_factor, name = "CFR (%)")
      ) +
      ggplot2$scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
      ggplot2$labs(
        title = paste("Weekly Cholera Case Increase –", .x),
        x = "Date",
        y = "Weekly Increase in Cases",
        caption ="
        ● Orange Dot: Watch alert
        ● Red Dot: Warning alert
        ● Green line: CERF allocation date
        ● Maroon dashed line: CFR"
      ) +
      gghdx$gghdx() +
      ggplot2$theme(
        plot.title = ggplot2$element_text(face = "bold", size = 14),
        axis.title = ggplot2$element_text(size = 11),
        axis.text = ggplot2$element_text(size = 10),
        plot.caption = ggplot2$element_text(size = 9, hjust = 0),
        panel.grid.minor = ggplot2$element_blank()
      )
  })
})
```


